---
output:
  rmarkdown::github_document:
    html_preview: no
---

<!--- README.md is generated from README.Rmd. Please edit that file -->

# Habitat classification data from Jung *et al.* 2020 (100 m global)

[![lifecycle](https://img.shields.io/badge/Lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![DOI](https://zenodo.org/badge/TODO.svg)](https://zenodo.org/badge/latestdoi/TODO)

```{r, include = FALSE}
knitr::opts_chunk$set(fig.path = "figures/README-", fig.align = "center")
```

This repository contains code to generate a global habitat classification data (based on Jung *et al.* 2020). The habitat classification data are provided at two thematic levels (i.e., level 1 and level 2). Specifically, source data are downloaded from the [Zenodo Digital Archive](https://doi.org/10.5281/zenodo.4058356), and then projected to the World Behrmann coordinate reference system using (100 m resolution; [ESRI:4326 coordinate reference system](https://epsg.io/54017)). A processed version of the data can be downloaded from the [Zenodo Digital Repository](https://zenodo.org/badge/latestdoi/TODO). **All credit should go to Jung *et al.* (2020), please cite their work if you use this dataset.**

```{r "functions", include = FALSE}
read_clr <- function(x) {
  txt <- readLines(x)
  split <- strsplit(txt, " ", fixed = TRUE)
  split <- split[!lengths(split) == 0]
  tibble::tibble(
    value = vapply(split, `[[`, character(1), 1),
    red = vapply(split, `[[`, character(1), 2),
    green = vapply(split, `[[`, character(1), 3),
    blue = vapply(split, `[[`, character(1), 4),
    alpha = vapply(split, `[[`, character(1), 5),
    name = vapply(split, FUN.VALUE = character(1), function(x) {
      paste(x[seq(6, length(x))], collapse = " ")
    }),
    color = rgb(red, green, blue, alpha, maxColorValue = 255)
  )
}
```

```{r "preliminary_processing", echo = FALSE, message = FALSE, warning = FALSE}
# load packages
library(dplyr)
library(terra)
library(ggplot2)

# find file paths
lvl1_path <- dir("results", "^.*lvl1.*tif$", full.names = TRUE)
lvl2_path <- dir("results", "^.*lvl2.*tif$", full.names = TRUE)

# find legend data
lvl1_metadata <- read_clr("data/level1.clr")
lvl2_metadata <- read_clr("data/level2.clr")

# load raster
lvl1_raster <- terra::rast(lvl1_path)
lvl2_raster <- terra::rast(lvl2_path)

# extract data to plot (note that we spatially sample data to reduce run time)
lvl1_plot_data <-
  lvl1_raster %>%
  terra::spatSample(size = 400000, method = "regular", as.raster = TRUE) %>%
  terra::as.data.frame(xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  dplyr::mutate(title = "Habitat classification (level 1)") %>%
  dplyr::mutate(subtitle = basename(lvl1_path)) %>%
  dplyr::mutate(value = as.character(value)) %>%
  dplyr::left_join(
    lvl1_metadata %>%
    dplyr::select(value, color, name),
    by = "value"
  )

lvl2_plot_data <-
  lvl2_raster %>%
  terra::spatSample(size = 400000, method = "regular", as.raster = TRUE) %>%
  terra::as.data.frame(xy = TRUE) %>%
  dplyr::rename(value = 3) %>%
  dplyr::mutate(title = "Habitat classification (level 2)") %>%
  dplyr::mutate(subtitle = basename(lvl2_path)) %>%
  dplyr::mutate(value = as.character(value)) %>%
  dplyr::left_join(
    lvl2_metadata %>%
    dplyr::select(value, color, name),
    by = "value"
  )

# create an underlying map
p <-
  ggplot() +
  coord_fixed(expand = FALSE) +
  facet_wrap(~ title + subtitle) +
  theme(
    legend.pos = "hide",
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.line = element_blank(),
    axis.ticks.length = unit(0, "null"),
    axis.ticks.margin = unit(0, "null"),
    panel.border = element_blank(),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    legend.text = element_text(size = 8),
    legend.margin = unit(0, "null"),
    plot.margin = unit(c(0.001, 0.001, 0.001, 0.01), "pt"),
    strip.background = ggplot2::element_rect(color = "black", fill = "black"),
    strip.text = ggplot2::element_text(color = "white")
  )

# make plots
lvl1_plot <-
  p +
  geom_tile(
    data = lvl1_plot_data,
    mapping = aes(x = x, y = y, fill = name)
  ) +
  scale_fill_manual(
    name = "Class",
    values = setNames(lvl1_metadata$color, lvl1_metadata$name)
  )

lvl2_plot <-
  p +
  geom_tile(
    data = lvl2_plot_data,
    mapping = aes(x = x, y = y, fill = name),
  ) +
  scale_fill_manual(
    name = "Class",
    values = setNames(lvl2_metadata$color, lvl2_metadata$name)
  )
```

```{r "lvl1_map", echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2.4, fig.width = 5.0, dpi = 250}
# render plot
print(lvl1_plot)
```

</br>

```{r "lvl2_map", echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2.4, fig.width = 5.0, dpi = 250}
# render plot
print(lvl2_plot)
```

## Metadata

The habitat classification data are stored in raster (GeoTIFF) files. The level 1 and level 2 data are stored separately. The file names contain the version number of the original datasets (i.e., per the digital object identifier). The latest version of the files are `r lvl1_path`{.R} and `r lvl2_path`{.R} (for level 1 and level 2 data, respectively).

The metadata for the `r basename(lvl1_path)`{.R} raster file is shown below.

```{r, echo = FALSE, comment = NA}
withr::with_dir("results", {
  cat(
    system(paste("gdalinfo", basename(lvl1_path)), intern = TRUE),
    sep = "\n"
  )
})
```

The metadata for the `r basename(lvl2_path)`{.R} raster file is shown below.

```{r, echo = FALSE, comment = NA}
withr::with_dir("results", {
  cat(
    system(paste("gdalinfo", basename(lvl2_path)), intern = TRUE),
    sep = "\n"
  )
})
```

## Usage

You can run the code to produce the dataset using the system command `make clean all`. After processing has successfully completed, the code will produce two raster (GeoTIFF) files in the `results` directory.

## Software requirements

Operating system

* Ubuntu (21.10) (300 Gb disk space available)

Software

* GNU make (version 4.1)
* GDAL (version 3.2.2)
* R (version 4.1.1)

System packages

* `gdal-bin`
* `libudunits2-dev`
* `libgdal-dev`
* `libgeos-dev`
* `libproj-dev`
* `libxml2-dev`
* `libssl-dev`

## Citation

This dataset was derived from Jung *et al.* (2020). If you use this dataset, you should cite their work - they deserve all the credit!

> Jung M, Dahal PR, Butchart SHM, Donald PF, De Lamo X, Lesiv M, Kapos V, Rondinini C, and Visconti P (2020a) A global map of terrestrial habitat types. *Scientific Data*, **7**:1--8. Available at https://doi.org/10.1038/s41597-020-00599-8.

> Jung M, Dahal PR, Butchart SHM, Donald PF, De Lamo X, Lesiv M, Kapos V, Rondinini C, and Visconti P (2020b) A global map of terrestrial habitat types (insert version) [Data set]. *Zenodo Digital Repository*. Available at https://doi.org/10.5281/zenodo.4058819.
